---
title: "Visualizing Cosponsorship in the Senate"
author: "Rashan Jibowu"
date: "08/08/2015"
output:
  html_document:
    keep_md: yes
---

```{r}

library(plyr)

```

Analysis Setup

```{r, cache=TRUE}

basePath <- c("./data/cosponsorship2010/")

# set up paths to the data
path.senate_bills <- paste0(basePath, "senate_bills.txt")
path.senate_members <- paste0(basePath, "senate_members/110_senators.txt")
path.senate_matrix <- paste0(basePath, "senate_matrices/110_senmatrix.txt")

# load the data
bills <- read.csv(path.senate_bills, header = FALSE)
senators <- read.csv(path.senate_members, header = FALSE)
cosponsorship <- read.csv(path.senate_matrix, header = FALSE)

str(bills)
str(senators)
#str(cosponsorship)

```

Clean data

Since there is so much data, let's focus on Senate Bills (`SN`) and ignore all other bill types

```{r clean data}

# Make the column names for senate bills human readable
colnames(bills) <- c("CongressNum", "BillType", "BillNum", "IsPrivate")
head(bills)

# subset for the 110th Congress
rBills <- bills[bills$CongressNum == 110,]

# Make the column names for senate members human readable
colnames(senators) <- c("Name", "THOMAS_ID", "ICPSR_ID")
head(senators)

# create unique bill Ids
billID <- paste0(rBills$BillType, rBills$BillNum);

# Each column in the matrix represents a bill facing the senate
colnames(cosponsorship) <- billID

# Each row in the matrix represents a senator
rownames(cosponsorship) <- senators[,c("Name")]

# test out the structure of the cosponsorship data frame
cosponsorship[1:5,1:5]

```


Make sense of the data

According to the codebook, the cosponsorship data can take one of several values:

Value | Meaning
------|--------
`1` | Congressman `i` sponsored bill `j`
`2` | Congressman `i` cosponsored bill `j`
`3` | Congressman `i` cosponsored bill `j` after withdrawing the previous cosponsorship from bill `j`
`5` | Congressman `i` withdrew a cosponsorship from bill `j`

So, to identify those bills with cosponsorship, we are looking for values 1, 2 and 3.
Values of 1 help us identify the original sponsor
Values of 2 and 3 help us identify the co-sponsors

This forms the basis of a social graph where senators are nodes and the bills that they have jointly sponsored represent the edges

First, let's find all bills that have cosponsorship

```{r}

# senate bills only
snOnly <- grep("^(SN)", colnames(cosponsorship), value = FALSE)
cosponsorship <- cosponsorship[,snOnly]

# remove columns (bills) that don't have cosponsorship
nocosponsor <- (colSums(cosponsorship) <= 1)

filtered <- cosponsorship[,!nocosponsor]
dim(filtered)

# if a column has only 1's, 0's and 5's, we should remove it because all cosponsors dropped out

noalldropout <- logical()

for (i in 1:ncol(filtered)) {
    noalldropout[i] <- any(filtered[,i] %in% c(2,3))
}

head(noalldropout)

filtered2 <- filtered[,noalldropout]

dim(filtered2)

```

Now, let's remove any senators without any activity

```{r}

lazysenator <- rowSums(filtered2) < 1

filtered2[lazysenator, 1]

```

Finally, let's write the data so that it can be visualized in D3.

```{r}

write.csv(filtered2, "./data/plotdata.csv")

```
